---
title: "PNW Flight Routes: A Flow Map"
subtitle: "Where does the Pacific Northwest fly?"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
```

```{r flow-map, fig.width=11, fig.height=6.5, fig.alt="Arc flow map of domestic flights departing PDX and SEA in 2014, overlaid on a dark US map"}
library(tidyverse)
library(pnwflights14)
library(maps)
library(geosphere)

data("flights",  package = "pnwflights14")
data("airports", package = "pnwflights14")

coords <- airports %>%
  select(faa, lat, lon) %>%
  filter(!is.na(lat), !is.na(lon))

route_counts <- flights %>%
  count(origin, dest, name = "n_flights")

routes <- route_counts %>%
  left_join(coords, by = c("origin" = "faa")) %>%
  rename(orig_lat = lat, orig_lon = lon) %>%
  left_join(coords, by = c("dest" = "faa")) %>%
  rename(dest_lat = lat, dest_lon = lon) %>%
  filter(
    !is.na(orig_lat), !is.na(dest_lat),
    dest_lon > -130, dest_lon < -65,
    dest_lat > 20,   dest_lat < 52
  )

make_arc <- function(orig_lon, orig_lat, dest_lon, dest_lat, n_flights, route_id) {
  pts <- gcIntermediate(
    c(orig_lon, orig_lat), c(dest_lon, dest_lat),
    n = 80, addStartEnd = TRUE
  )
  tibble(lon = pts[, 1], lat = pts[, 2],
         n_flights = n_flights, route_id = route_id)
}

arcs <- pmap_dfr(
  list(
    orig_lon  = routes$orig_lon,  orig_lat  = routes$orig_lat,
    dest_lon  = routes$dest_lon,  dest_lat  = routes$dest_lat,
    n_flights = routes$n_flights, route_id  = seq_len(nrow(routes))
  ),
  make_arc
)

us_states <- map_data("state")
pnw       <- coords %>% filter(faa %in% c("PDX", "SEA"))

top_dests <- route_counts %>%
  group_by(dest) %>%
  summarise(total = sum(n_flights), .groups = "drop") %>%
  slice_max(total, n = 25) %>%
  left_join(coords, by = c("dest" = "faa")) %>%
  filter(!is.na(lat), lon > -130, lon < -65, lat > 20, lat < 52)

ggplot() +
  geom_polygon(
    data = us_states,
    aes(x = long, y = lat, group = group),
    fill = "#1c2d3a", color = "#2e4a5e", linewidth = 0.15
  ) +
  geom_path(
    data = arcs %>% arrange(n_flights),
    aes(x = lon, y = lat, group = route_id,
        alpha = n_flights, linewidth = n_flights),
    color = "#f4a261"
  ) +
  scale_alpha_continuous(range = c(0.05, 0.72), guide = "none") +
  scale_linewidth_continuous(range = c(0.10, 1.7), guide = "none") +
  geom_point(
    data = top_dests,
    aes(x = lon, y = lat),
    color = "white", size = 1.3, alpha = 0.5
  ) +
  geom_point(
    data = pnw,
    aes(x = lon, y = lat),
    color = "#e9c46a", size = 5, shape = 18
  ) +
  geom_label(
    data = pnw %>% filter(faa == "PDX"),
    aes(x = lon, y = lat, label = faa),
    color = "#e9c46a", fill = "#0d1b2a",
    size = 3.5, fontface = "bold",
    label.padding = unit(0.15, "lines"),
    nudge_x = -2.5, nudge_y = 0
  ) +
  geom_label(
    data = pnw %>% filter(faa == "SEA"),
    aes(x = lon, y = lat, label = faa),
    color = "#e9c46a", fill = "#0d1b2a",
    size = 3.5, fontface = "bold",
    label.padding = unit(0.15, "lines"),
    nudge_y = 1.6
  ) +
  coord_fixed(ratio = 1.3, xlim = c(-125, -66), ylim = c(24, 50)) +
  labs(
    title    = "PNW Flight Routes (2014)",
    subtitle = "All domestic routes departing Portland (PDX) and Seattle (SEA)\nArc thickness and opacity scaled with number of scheduled flights",
    caption  = "Source: pnwflights14"
  ) +
  theme_void() +
  theme(
    plot.background  = element_rect(fill = "#0d1b2a", color = NA),
    panel.background = element_rect(fill = "#0d1b2a", color = NA),
    plot.title    = element_text(color = "#ffffff", size = 20, face = "bold",
                                 margin = margin(t = 15, l = 20)),
    plot.subtitle = element_text(color = "#adb5bd", size = 11,
                                 margin = margin(t = 6, l = 20, b = 8)),
    plot.caption  = element_text(color = "#6c757d", size = 8,
                                 margin = margin(b = 12, r = 15), hjust = 1),
    plot.margin   = margin(10, 15, 10, 15)
  )
```

---

## What Kind of Chart Is This?

A **flow map** (also called a *connection map* or *arc map*) is a thematic map that draws lines, often curved arcs, between pairs of geographic locations to represent movement, migration, or communication between them.

| Element | Role |
|---------|------|
| **Origin node** | Where flow begins (here: PDX, SEA) |
| **Destination node** | Where flow ends (US airports) |
| **Arc** | The connection; visual weight encodes quantity |

Flow maps date back to around 1838, designed by Irish engineer Henry Drury Harness to depict cargo flow across Ireland. Modern implementations use great-circle routes, the shortest path between two points on a sphere, rather than straight lines, which matters especially for long-haul routes.

---

## The Data

The data come from the [`pnwflights14`](https://github.com/ismayc/pnwflights14) R package, covering **Pacific Northwest departures in calendar year 2014**.

Three tables are used:

- `flights`: ~162,000 rows, one per scheduled departure. Key columns: `origin`, `dest`, `carrier`, `dep_delay`, `distance`.
- `airports`: one row per airport with FAA code, name, and decimal lat/lon.
- `airlines`: lookup table mapping IATA carrier codes to full airline names.

**Quantitative variables:** number of flights per route, great-circle arc distance.
**Qualitative variables:** origin airport (PDX vs. SEA), destination airport code, US state geography.

---

## What Is the Chart Showing?

The map answers one question: **how is the Pacific Northwest connected to the rest of the country by air?**

Key things a viewer can read:

- **Dominant routes**: thick, bright arcs (PDX/SEA -> LAX, SFO, ORD) are immediately visible.
- **Geographic reach**: the fan of arcs shows that PNW flyers can reach both coasts directly, though the West Coast dominates in frequency.
- **PDX vs. SEA differences**: the two origin diamonds allow a reader to trace each airport's routes semi-independently.
- **Hub-and-spoke structure**: dense arcs toward a few massive hubs (Chicago, San Francisco, Los Angeles) show how most airlines route passengers through central hubs rather than point-to-point, with notable exceptions like Southwest and Alaska who also serve many secondary airports directly.

Flow maps are intentionally designed for the casual observer. For precise counts, a table or bar chart is superior. Their strength is their accessibility: they communicate the shape of a dataset immediately to non-domain experts, which is where I find them most valuable.
